Text-mining clinically relevant cancer biomarkers for curation into the CIViC database

### 2018-12-18 paper/paper.Rmd
Minor tweaks to paper

## Abstract
Precision oncology involves analysis of individual cancer samples to understand the genes and pathways involved in the development and progression of a cancer. To improve patient care, knowledge of diagnostic, prognostic, predisposing and drug response markers is essential. Several knowledgebases have been created by different groups to collate evidence for these associations. These include the open-access Clinical Interpretation of Variants in Cancer (CIViC) knowledgebase. These databases rely on very time-consuming manual curation from highly-skilled experts who read and interpret the relevant biomedical literature. To aid in this curation and provide the greatest coverage for these databases, particularly CIViC, we propose the use of text mining approaches to extract these clinically relevant biomarkers from all available published literature. A group of cancer genomics experts annotated biomarkers and their clinical associations discussed in 800 sentences and achieved good inter-annotator agreement. We used a supervised learning approach to construct the CIViCmine knowledgebase (http://bionlp.bcgsc.ca/civicmine/) and extracted `r paper.sentenceCount` relevant sentences from PubMed abstracts and Pubmed Central Open Access full text papers. CIViCmine contains over `r paper.biomarkerCount` biomarkers associated with `r paper.geneCount` genes, `r paper.drugCount` drugs and `r paper.cancerCount` cancer types, representing `r paper.abstractCount` abstracts and `r paper.articleCount` full-text publications. Through integration with CIVIC, we provide a prioritised list of curatable biomarkers as well as a resource that is valuable to other knowledgebases and precision cancer analysts in general.

@@

The ability to stratify patients into groups that are clinically related is an important step towards a personalized approach to cancer. Over time, a growing number of biomarkers have been developed in order to select patients who are more likely to respond to certain treatments. These biomarkers have also been valuable for prognostic purposes and for understanding the underlying disease biology by defining different molecular subtypes of cancers that should be treated in different ways (e.g. *ERBB2*/*ESR1*/*PGR* testing in breast cancer [@onitilo2009breast]). Immunohistochemistry techniques are a primary approach for testing samples for diagnostic markers (e.g. CD15 and CD30 for Hodgkin’s disease [@rudiger1998differential]). Recently, the lower cost and increasing speed of sequencing has also allowed the DNA and RNA of individual patient samples to be characterized for clinical applications [@prasad2016precision]. Throughout the world, this technology is beginning to inform clinician decisions on which treatments to use [@shrager2014rapid]. Such efforts are dependent on comprehensive and current understanding of the clinical relevance of variants. For example, the Personalized Oncogenomics project at BC Cancer identifies somatic events in the genome such as point mutations, copy number variations and large structural changes and, in conjunction with gene expression data, generates a clinical report to provide an ‘omic picture of a patient’s tumor [@laskin2015lessons].

The high genomic variability observed in cancers means that each patient sample includes a large number of new mutations, many of which have never been documented before [@chang2016identifying]. The phenotypic impact of most of these mutations is difficult to discern. This problem is exacerbated by the driver/passenger mutation paradigm where only a fraction of mutations are essential to the cancer (drivers) while many others have occurred through mutational processes that are irrelevant to the cancer (passengers). An analyst trying to understand a new patient sample typically performs a literature review for each gene and specific variant. This is needed to understand its relevance in a cancer type, characterize the driver/passenger role of its observed mutations, and gauge the relevance for clinical decision making.

Several groups have built their own in-house knowledgebases, which are developed as analysts examine increasing numbers of cancer patient samples. This tedious and largely redundant effort represents a substantial interpretation bottleneck impeding the progress of precision medicine [@good2014organizing]. To encourage a collaborative effort, the CIViC knowledgebase (https://civicdb.org) was launched to provide a wiki-like, editable online resource where community contributed edits and additions are moderated by experts in order to maintain high quality [@griffith2017civic]. The resource provides information about clinically-relevant variants in cancer described in the peer-reviewed literature. Variants include protein-coding point mutations, copy number variations, epigenetic marks, gene fusions, aberrant expression levels and other ‘omic events. It supports four types of evidence associating biomarkers with different classes of clinical relevance (also known as evidence types).

@@

Several other groups have created knowledgebases to aid clinical interpretation of cancer genomes. Many of these projects have joined the Variant Interpretation for Cancer Consortium (VICC, http://cancervariants.org/) to coordinate these efforts and have created a federated search mechanism to allow easier analysis across multiple knowledgebases [@wagner2018harmonized]. The CIViC project is co-leading this effort along with OncoKB [@chakravarty2017oncokb], the Cancer Genome Interpreter [@tamborero2018cancer], Precision Medicine Knowledge base [@huang2017cancer], Molecular Match, JAX-Clinical Knowledge base [@patterson2016clinical] and others.

@@

The literature on clinically relevant cancer mutations is growing at an extraordinary rate. For instance there were only 5 publications with BRAF V600E in title or abstract in PubMed in 2004 compared to 454 papers in 2017. In order to maintain a high quality and up-to-date knowledge-base, a curation pipeline must be established. This typically involves a queue for papers, triaging those that should be curated and then assignment to a highly experienced curator. This prioritisation step is important given the limited time of curators and the potentially vast number of papers to be reviewed. Prioritisation must identify papers that contain knowledge that is of current relevance to users of the knowledgebase. For instance, selecting papers for drugs that are no longer clinically approved would not be valuable to the knowledgebase.

>>>>

## Abstract
Precision oncology involves analysis of individual cancer samples to understand the genes and pathways involved in the development and progression of a cancer. To improve patient care, knowledge of diagnostic, prognostic, predisposing and drug response markers is essential. Several knowledgebases have been created by different groups to collate evidence for these associations. These include the open-access Clinical Interpretation of Variants in Cancer (CIViC) knowledgebase. These databases rely on time-consuming manual curation from skilled experts who read and interpret the relevant biomedical literature. To aid in this curation and provide the greatest coverage for these databases, particularly CIViC, we propose the use of text mining approaches to extract these clinically relevant biomarkers from all available published literature. To this end, a group of cancer genomics experts annotated biomarkers and their clinical associations discussed in 800 sentences and achieved good inter-annotator agreement. We then used a supervised learning approach to construct the CIViCmine knowledgebase (http://bionlp.bcgsc.ca/civicmine/), extracting `r paper.sentenceCount` relevant sentences from PubMed abstracts and Pubmed Central Open Access full text papers. CIViCmine contains over `r paper.biomarkerCount` biomarkers associated with `r paper.geneCount` genes, `r paper.drugCount` drugs and `r paper.cancerCount` cancer types, representing `r paper.abstractCount` abstracts and `r paper.articleCount` full-text publications. Through integration with CIVIC, we provide a prioritised list of curatable biomarkers as well as a resource that is valuable to other knowledgebases and precision cancer analysts in general.

@@

The ability to stratify patients into groups that are clinically related is an important step towards a personalized approach to cancer. Over time, a growing number of biomarkers have been developed in order to select patients who are more likely to respond to certain treatments. These biomarkers have also been valuable for prognostic purposes and for understanding the underlying biology of the disease by defining different molecular subtypes of cancers that should be treated in different ways (e.g. *ERBB2*/*ESR1*/*PGR* testing in breast cancer [@onitilo2009breast]). Immunohistochemistry techniques are a primary approach for testing samples for diagnostic markers (e.g. CD15 and CD30 for Hodgkin’s disease [@rudiger1998differential]). Recently, the lower cost and increasing speed of geneome sequencing has also allowed the DNA and RNA of individual patient samples to be characterized for clinical applications [@prasad2016precision]. Throughout the world, this technology is beginning to inform clinician decisions on which treatments to use [@shrager2014rapid]. Such efforts are dependent on comprehensive and current understanding of the clinical relevance of variants. For example, the Personalized Oncogenomics project at BC Cancer identifies somatic events in the genome such as point mutations, copy number variations and large structural changes and, in conjunction with gene expression data, generates a clinical report to provide an ‘omic picture of a patient’s tumor [@laskin2015lessons].

The high genomic variability observed in cancers means that each patient sample includes a large number of new mutations, many of which may have never been documented before [@chang2016identifying]. The phenotypic impact of most of these mutations is difficult to discern. This problem is exacerbated by the driver/passenger mutation paradigm where only a fraction of mutations are essential to the cancer (drivers) while many others have occurred through mutational processes that are irrelevant to the progression of the disease (passengers). An analyst trying to understand a patient sample typically performs a literature review for each gene and specific variant which is needed to understand its relevance in a cancer type, characterize the driver/passenger role of its observed mutations, and gauge the relevance for clinical decision making.

A number of groups have built in-house knowledgebases, which are developed as analysts examine increasing numbers of cancer patient samples. This tedious and largely redundant effort represents a substantial interpretation bottleneck impeding the progress of precision medicine [@good2014organizing]. To encourage a collaborative effort, the CIViC knowledgebase (https://civicdb.org) was launched to provide a wiki-like, editable online resource where community contributed edits and additions are moderated by experts in order to maintain high quality variant curation [@griffith2017civic]. The resource provides information about clinically-relevant variants in cancer described in the peer-reviewed literature. Variants include protein-coding point mutations, copy number variations, epigenetic marks, gene fusions, aberrant expression levels and other ‘omic events. It supports four types of evidence associating biomarkers with different classes of clinical relevance (also known as evidence types).

@@

Several groups have created knowledgebases to aid clinical interpretation of cancer genomes, many of whom have joined the Variant Interpretation for Cancer Consortium (VICC, http://cancervariants.org/) . VICC is an initiative that aims to coordinate variant interpretation efforts and, to this end, has created a federated search mechanism to allow easier analysis across multiple knowledgebases [@wagner2018harmonized]. The CIViC project is co-leading this effort along with OncoKB [@chakravarty2017oncokb], the Cancer Genome Interpreter [@tamborero2018cancer], Precision Medicine Knowledge base [@huang2017cancer], Molecular Match, JAX-Clinical Knowledge base [@patterson2016clinical] and others.

@@

The literature on clinically relevant cancer mutations is growing at an extraordinary rate. For instance, only 5 publications in PubMed mentioned BRAF V600E in the title or abstract in 2004 compared to 454 papers in 2017. In order to maintain a high quality and up-to-date knowledge-base, a curation pipeline must be established. This typically involves a queue for papers, triaging those that should be curated and then assignment to a highly experienced curator. This prioritisation step is important given the limited time of curators and the potentially vast number of papers to be reviewed. Prioritisation must identify papers that contain knowledge that is of current relevance to users of the knowledgebase. For instance, selecting papers for drugs that are no longer clinically approved would not be valuable to the knowledgebase.
